---
output: github_document
---

[![Travis build status](https://travis-ci.org/mathesong/kinfitr.svg?branch=master)](https://travis-ci.org/mathesong/kinfitr)
[![Coverage Status](https://img.shields.io/codecov/c/github/mathesong/kinfitr/master.svg)](https://codecov.io/github/mathesong/kinfitr?branch=master)
  

# kinfitr : PET Kinetic Modelling using R

## Overview

kinfitr is a package for PET Kinetic Modelling Using R. The main goal of this package is to make it easier to perform more reproducible research in PET research. Furthermore, both kinfitr and the language it is written in, [R](https://cran.r-project.org), are free to download. Furthermore, the primary IDE for R, [RStudio](https://www.rstudio.com/), now comes with [R Notebooks](http://rmarkdown.rstudio.com/r_notebooks.html), which makes it particularly easy to generate reproducible reports.


## Installation

This package is currently only available on GitHub.  It can be downloaded as follows:

```{r, eval = FALSE}
# install.packages("devtools")  # If you do not already have devtools
devtools::install_github("mathesong/kinfitr")
```

There is also a docker container if you would prefer not to have to download everything. If you download Docker, you can pull the container and start it with the following commands:

```{}
docker pull mathesong/kinfitr_docker
docker run -e PASSWORD=XYZ --name rstudio -p 8787:8787 mathesong/kinfitr_docker
```

And the the rstudio session can be opened in a browser pointing to `http://localhost:8787`.

## Usage

### Data Structure

The optimal data structure for _kinfitr_ is that of a nested tibble, with rows representing the desired level of chunking, e.g. whether modelling be across individuals, or across ROIs within individuals). The package contains two datasets: `pbr28` containing [$^{11}$C]PBR28 TACs for testing models involving arterial input function, and `simref` containing simulated TACs of a tracer with a reference region, for testing reference tissue models.


```{r, message=F}
library(tidyverse)
library(kinfitr)
library(knitr)

data(simref)
```

Thus the data looks as follows:

```{r}
head(simref)
```

...and inside each nested tibble is the following:

```{r}
head(simref$tacs[[1]])
```

### Fitting a Model for a single TAC

As a conscious decision, almost all input arguments of times or radioactivity concentrations are as numeric vectors. This allows the functions to be used at any stage of an analysis, and does not require complicated data structures created in earlier steps. So let's create vectors and run a model.

```{r srtmfit}
times <- simref$tacs[[1]]$Times
tac <- simref$tacs[[1]]$ROI1
reference <- simref$tacs[[1]]$Reference
weights <- simref$tacs[[1]]$Weights

srtmfit <- srtm(t_tac = times, reftac = reference,
      roitac = tac,weights = weights)

plot_kinfit(srtmfit)
```



### Fitting a Model to Many TACs

For many TACs, we can use the fabulous `purrr` package to iterate through our nested tibble.  Let's also examine how we might decide to chunk the data, using MRTM1 and MRTM2.

What we want to do from here is to model the data using MRTM1 and MRTM2.  Our plan is as follows:

* MRTM1 fits BP~ND~ and k2'
    
* MRTM2 fits BP~ND~ (using k2' from MRTM1 from a high-binding region)

So what we'll do:

* Fit MRTM1 to one region of each PET Measurement
    
* Fit MRTM2 to all regions of each PET Measurement



#### Fitting k2prime using MRTM1 using purrr::map

```{r mrtm2fit}
simref <- simref %>%
  
  group_by(PET) %>%     # Group by each PET measurement
  
  mutate(mrtm1_fit = map(tacs, ~mrtm1(t_tac = .x$Times, reftac = .x$Reference,      # Add MRTM1 fit column
                                      roitac = .x$ROI1, weights = .x$Weights))) %>%
  
  mutate(k2prime = map_dbl(mrtm1_fit, c('par', 'k2prime')))     # Extract k2prime from the fit output

plot_kinfit(simref$mrtm1_fit[[1]])     # Plot the first TAC
```

Now we have a k2' value for each individual from a single region (ROI1).

```{r}
head(simref)
```


#### Tidy Data: Gathering into Long format

Now we want to use the k2prime from fitting MRTM1 to one region from each PET measurement.  Now, we want to chunk the data a little bit differently: we want to make the arrangement a little bit longer: each TAC which we wish to model should be a row.

```{r}
long_simref <- simref %>%
  
  select(-mrtm1_fit) %>% # Remove the fit object
  
  unnest() %>%    # Unnest
  
  select(-StartTime, -Duration) %>% 
  
  gather(Region, TAC, -PET, -Subjname, -PETNo, -Weights, 
         -Times, -k2prime, -Reference) %>%    # Gather the data into long-er format
  
  group_by(PET, Subjname, PETNo, Region, k2prime) %>%    # Group by more than just PET
  
  nest(.key=tacs) %>%       # Nest the data again %>% 

  arrange(Subjname, Region)
```

This produces data which looks like this:

```{r}
head(long_simref)
```


For which the TACs nested object looks like this:

```{r}
head( long_simref$tacs[[1]] )
```


#### Fitting MRTM2 using purrr::pmap

Now we can iterate through this list, using the fitted k2prime. We can't use `map()` any longer, as we'll be iterating through 2 lists.  For this, we could either use `map2()`, or we can use `pmap()`, which allows us to iterate through an unlimited numberof columns.  Let's go with the latter for now to show how we could do this.

First we define a function for the iteration:

```{r}
dofit_mrtm2 <- function(tacs, k2prime) {
  mrtm2(t_tac = tacs$Times, reftac = tacs$Reference, 
            roitac = tacs$TAC, weights = tacs$Weights,
            k2prime = k2prime)
}
```

... and then we apply it:

```{r}
long_simref <- long_simref %>%

  mutate(fit_mrtm2 = pmap(list(tacs, k2prime), dofit_mrtm2)) %>%

  mutate(bp_mrtm2 = map_dbl(fit_mrtm2, c('par', 'bp')))


plot_kinfit(long_simref$fit_mrtm2[[6]])
```


In this way, we can run several different models on the data, keep the fits in their own separate columns, and plot them at will.  And we can do all of this within the tidyverse paradigm.



## Implemented Models

__Reference Region Models__

* Simplified Reference Tissue Model (SRTM) _(Lammertsma & Hume, 1996)_

* Non-Invasive Logan Plot _(Logan et al., 1996)_

* Non-Invasive Multilinear Logan Plot _(Turkheimer et al., 2003)_

* Ichise Multilinear Reference Tissue Model 1 (MRTM1) _(Ichise et al., 2003)_

* Ichise Multilinear Reference Tissue Model 2 (MRTM2) _(Ichise et al., 2003)_

* Patlak Reference Tissue Model _(Patlak & Blasberg, 1985)_



__Models Requiring Arterial Input__

* One-Tissue Compartment Model

* Two-Tissue Compartment Model

* Logan Plot _(Logan et al., 1990)_

* Multilinear Logan Plot _(Turkheimer et al., 2003)_

* Ichise Multilinear Analysis 1 (MA1) _(Ichise et al., 2002)_

* Ichise Multilinear Analysis 2 (MA2) _(Ichise et al., 2002)_

* Patlak Plot _(Patlak et al., 1983)_



__Other Models__

* Simultaneous Estimation of Non-Displaceable Binding (SIME) _(Ogden et al., 2015)_





## To-Do

__General__
    
* Vignettes
    
__Additions__

    

__Improvements__

* Tidy up

    * Documentation still quite rough and several inconsistencies
  
    * T-star finders have lots of code duplication: should be more generic
    
* Revise vignette

* Update plyr functions to dplyr functions

* Get the package CRAN-ready
